Hydra Battles and Cie

### 2020-12-03 doc/chapter-powers.tex
work on addition chains chapter
add example on Fibonacci

Matrix exponentiation is a well-known method for computing Fibonacci numbers:
\index{Fibonacci numbers!Matrix exponentiation}

Import M2.

Arguments M2_mult {A} plus mult  _  _.
Arguments mat {A} _ _ _ _.
Arguments Id2 {A}  _ _.

Definition fibonacci (n:N) :=
 c00 N  (N_bpow  (M2_mult Nplus Nmult) 
                 (Id2  0%N 1%N)
                 (mat  1 1 1 0)%N 
                 n).

Compute fibonacci 20.
\end{Coqsrc}

\begin{Coqanswer}
= 10946%N
     : N  
\end{Coqanswer}

@@

\subsubsection{Remark}
\label{sect:faster}

Our function \texttt{N\_bpow} is really logarithmic. Let us make a comparative 
test with Standard Library's exponentiation function on type \texttt{N} (see section~\vref{sect:slow-computation}).

\begin{Coqsrc}
Time Compute (N_bpow N.mul 1 1 56666667)%N.  
\end{Coqsrc}

\begin{Coqanswer}
Finished transaction in 0. secs (0.u,0.s) (successful)  
\end{Coqanswer}

>>>>

\subsection{Computing Fibonacci Numbers}

The sequence of Fibonacci numbers is defined by the following equations:

\begin{align}
F_0 & = 1 \\
F_1 & = 1 \\
F_n & = F_{n-1} + F_{n-2} \quad (n \geq 2)
\end{align}

In \coq{}, one can define this function by simple recursion.

\emph{From Library
\href{../theories/html/hydras.additions.Fib2.html}{additions.Fib2}}

\begin{Coqsrc}
Fixpoint fib (n:nat) : N :=
  match n with
    0%nat | 1%nat => 1%N
  | S (S p as q) => fib p + fib q
  end.

Compute fib 20.
\end{Coqsrc}

\begin{Coqanswer}
= 10946 : N
\end{Coqanswer}

In~\cite{BC04}, several exercices~\footnote{Exercises 9.8 (page 270), 9.10 (page 271), 9.15 (page 276), 9.17 (page 284), and 15.8 (page 418).}  
present ways to compute Fibonacci numbers, with the less number of recursive calls  as possible. Please note that these optimizations and the formal proof of their correctness are \emph{ad-hoc}, \emph{i.e.} exclusively written for the 
Fibonacci numbers.
In contrast, the optimizations we present in this document apply, in their vast majority, \emph{generic} techniques of efficient computation of powers in a monoid. 
This example of Fibonacci numbers has been developed with Yves Bertot, who wrote a first version with \texttt{SSreflect/Mathcomp}~\cite{MCB, SSR}.

\subsubsection{Using 2x2 integer matrices}
\index{Fibonacci numbers!Matrix exponentiation}

The following properties are well known. They are left as an exercise, since they are not part of our development.

\index{Exercises}

\begin{exercise}
  \begin{enumerate}
  \item

Prove in \coq{} the following equality (for any $n\geq 2$). \label{fibmat-eq1}

\[
\left(
  \begin{array}{cc}
    1 & 1 \\
    1 & 0 
  \end{array}
\right)
\left(
  \begin{array}{cc}
    F_{n}& F_{n-1} \\
    F_{n-1} & F_{n-2}
  \end{array}
\right)
=
\left(
  \begin{array}{cc}
    F_{n+1}& F_{n} \\
    F_{n} & F_{n-1} 
  \end{array}
\right)
\]

\item Infer (still in \coq{}) the following equality (still for $n\geq 2$).

\[
\left(
  \begin{array}{cc}
    F_{n}& F_{n-1} \\
    F_{n-1} & F_{n-2} 
  \end{array}
\right)
= 
\left(
  \begin{array}{cc}
    1 & 1 \\
    1 & 0 
  \end{array}
\right)^n
\]

\item Write a function using the previous equality for computing the $n$-th Fibonacci number, and prove its equivalence with \texttt{fib}.

\end{enumerate}
\end{exercise}

\subsubsection{Removing duplicate computations}

Yves Bertot's optimization relies on the observation that all the powers of
\(  \left(
  \begin{array}{cc}
    1 & 1 \\
    1 & 0 
  \end{array}
\right) \) have the form 
\(  \left(
  \begin{array}{cc}
    a+b  & a \\
    a & b
  \end{array}
\right) \) where $a$ and $b$ are natural numbers.

Thus, it is possible to remove duplicate data and computations by reflecting matrix multiplication and identity into $\mathbb{N}\times\mathbb{N}$.

If we define $\varphi(a,b)  \left(
  \begin{array}{cc}
    a+b  & a \\
    a & b
  \end{array}
\right)$, then $\varphi(a,b)\times \varphi(c,d)=\varphi(ac + ad + bc, ac + bd)$, and
$\varphi(0,1)=  \left(
  \begin{array}{cc}
    1 & 0 \\
    0 & 1 
  \end{array}
\right) $.

\index{Exercises}
\begin{exercise}
  Prove formally these properties. \emph{Please note that their proof is not needed in our development, they just help to understand the following optimization.}
\end{exercise}

So, let us define a binary operation, which makes $\mathbb{N}\times\mathbb{N}$ a monoid (with $(0,1)$ as neutral element).

\emph{From Library
\href{../theories/html/hydras.additions.Fib2.html}{additions.Fib2}}

\begin{Coqsrc}
Definition mul2 (p q : N * N) :=
  match p, q with (a, b),(c,d) => (a*c + a*d + b*c, a*c + b*d) end.

Instance Mul2 : Monoid  mul2 (0,1).
(* Proof omitted *)
\end{Coqsrc}

The following lemma is a simplification of the equality~\vref{fibmat-eq1}.

\begin{Coqsrc}
Lemma next_fib (n:nat) : mul2 (1,0) (fib (S n), fib n) =
                         (fib (S (S n)), fib (S n)).  
\end{Coqsrc}

By induction\footnote{Simple induction, since $n$ has type \texttt{nat}.}
 over $n$, we obtain the following equality.

 \begin{Coqsrc}
Lemma fib_power n : fib n = let (a,b) := power (M:=Mul2) (1,0) n
                            in (a+b).   
\end{Coqsrc}

Thus, any function able to compute more or less efficiently powers in a monoid will
give an algorithm for computing Fibonacci numbers. Unlike the \emph{ad-hoc} aforementionned proofs of~\cite{BC04}, the correctness of such an algorithm is a direct consequence 
of the correctness of the used powering function.
Several examples will be presented in the rest of this document.

\begin{todo}
Put links when done.
\end{todo}

% \begin{Coqsrc}

% Import M2.

% Arguments M2_mult {A} plus mult  _  _.
% Arguments mat {A} _ _ _ _.
% Arguments Id2 {A}  _ _.

% Definition fibonacci (n:N) :=
%  c00 N  (N_bpow  (M2_mult Nplus Nmult) 
%                  (Id2  0%N 1%N)
%                  (mat  1 1 1 0)%N 
%                  n).

% Compute fibonacci 20.
% \end{Coqsrc}

% \begin{Coqanswer}
% = 10946%N
%      : N  
% \end{Coqanswer}

@@

% \subsubsection{Remark}
% \label{sect:faster}

% Our function \texttt{N\_bpow} is really logarithmic. Let us make a comparative 
% test with Standard Library's exponentiation function on type \texttt{N} (see section~\vref{sect:slow-computation}).

% \begin{Coqsrc}
% Time Compute (N_bpow N.mul 1 1 56666667)%N.  
% \end{Coqsrc}

% \begin{Coqanswer}
% Finished transaction in 0. secs (0.u,0.s) (successful)  
% \end{Coqanswer}
